---
title: "assignment2-karim"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=3)
```

Daan Moll (VU: 2559766 / UvA: 11929332)
Ramon Bussing (VU: 2687261 / UvA: 10719482)
Karim Semin (VU: ksn222 / UvA: 11285990)

Group number: 49

For all exercises, an $\alpha$ significance level of .05 is used, unless stated otherwise.

Exercise 1. Moldy bread
If left alone bread will become moldy, rot or decay otherwise. To investigate the influence of temperature and humidity on this process, the time to decay was measured for 18 slices of white bread, which were placed in 3 different environments and humidified or not. The data are given in the file bread.txt, with the first column time to decay in hours, the second column the environment (cold, warm or intermediate temperature) and the third column the humidity.
a) The 18 slices came from a single loaf, but were randomized to the 6 combinations of conditions. Present an R-code for this randomization process.

```{r}
bread = read.table("bread.txt",header=TRUE)
bread
```

The randomization process of the data is done as followed:
```{r}
I=3; J=2; N=3
rbind(rep(1:I,each=N*J),rep(1:J,N*I),sample(1:(N*I*J)))
```


b) Make two boxplots of hours versus the two factors and two interaction plots (keeping the two factors fixed in turn).
```{r}
attach(bread)
boxplot(hours~environment)
boxplot(hours~humidity)
```
```{r}
interaction.plot(environment, humidity, hours)
interaction.plot(humidity, environment, hours)
```
c) Perform an analysis of variance to test for effect of the factors temperature, humidity, and their interaction. Describe the interaction effect in words.
```{r}
bread$environment=as.factor(bread$environment)
bread$humidity=as.factor(bread$humidity)

breadov=lm(hours~environment*humidity)
anova(breadov)
# in ass 1 zeggen ze bij 5b:
# b) -0.4, check summary(chickaov) and estimate weights. Highest weight is the best--> sunflower
# dit is nu ook van teopassing denk ik?
```
The p-value for both  $H_0$ with factors $\alpha_i =0$ and $\beta_i =0$ for all $i$ is <0.001. Based 
The p-value for $H0 :\gamma_{i,j} =0$ :too is <0.001. This implies that we can not reject $H_0$, since there may be an interaction effect between the environment and humidity.

d) Which of the two factors has the greatest (numerical) influence on the decay? Is this a good question?
e) Check the model assumptions by using relevant diagnostic tools. Are there any outliers?

