---
title: "assignment1"
output: html_document
---

```{r setup, include=FALSE}
options(digits=3)
knitr::opts_chunk$set(echo = TRUE)
```

The data set birthweight.txt contains the birthweights of 188 newborn babies. We are interested in finding the underlying (population) mean μ of birthweights.

a) Check normality of the data. Compute a point estimate for μ. Derive, assuming normality (irrespective of your conclusion about normality od the data), a bounded 90% confidence interval for μ.

```{r echo=FALSE}
birth = read.table("birthweight.txt",header=TRUE)
birth$birthweight
mean(birth$birthweight)

shapiro.test(birth$birthweight)[2]
```

b) An expert claims that the mean birthweight is bigger than 2800, verify this claim by using a t-test. What is the outcome of the test if you take α = 0.1? And other values of α?

```{r}
alpha=0.1
t.test(birth$birthweight,mu=2800,alt="l",conf.level = 1-alpha)

alpha=0.05
t.test(birth$birthweight,mu=2800,alt="l",conf.level = 1-alpha)

alpha=0.01
t.test(birth$birthweight,mu=2800,alt="l",conf.level = 1-alpha)
```

c) In the R-output of the test from b), also a confidence interval is given, but why is it different from the confidence interval found in a) and why is it one-sided?
1 sided because you check if it is greater than mu and not both lower and greater.

Different because ... The confidence interval is one-sided because we only look if the mu is greater than 2800.





