---
title: "assignment1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The data set birthweight.txt contains the birthweights of 188 newborn babies. We are interested in finding the underlying (population) mean μ of birthweights.

a) Check normality of the data. Compute a point estimate for μ. Derive, assuming normality (irrespective of your conclusion about normality od the data), a bounded 90% confidence interval for μ.

```{r echo=FALSE}
birth = read.table("birthweight.txt",header=TRUE)
birthweight = birth$birthweight
```

Point estimate for μ:
```{r}
mean(birthweight)
```

Check normality with Shapiro-Wilk test:
```{r}
shapiro.test(birthweight)[2]
```
p is non-significant. This implies that h0 is accepted (the data is normally distributed).


b) An expert claims that the mean birthweight is bigger than 2800, verify this claim by using a t-test. What is the outcome of the test if you take α = 0.1? And other values of α?

If we test this hypothesis with α = 0.1 we get:
```{r}
alpha = .1
t.test(birthweight, alternative = "l", conf.level = 1-alpha, mu = 2800)[3]
```

p = .9864, therefore, h0 is accepted. If we try this for alpha = .05, .01, .001, we get the following p values:
```{r}
alpha = .05
t.test(birthweight, alternative = "l", conf.level = 1-alpha, mu = 2800)[3]

alpha = .05

```



c) In the R-output of the test from b), also a confidence interval is given, but why is it different from the confidence interval found in a) and why is it one-sided?
