---
title: "exercise2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
A telecommunication company has entered the market for mobile phones in a new country. The company’s marketing manager conducts a survey of 200 new subscribers for mobile phones. The results of the survey are in the data set telephone.txt, which contains the first month bills X1 , . . . X200 , in euros.


a) Make an appropriate plot of this data set. What marketing advice(s) would you give to the marketing manager? Are there any inconsistencies in the data? If so, try to fix these.

```{r}
telephone_data = read.table("telephone.txt",header=TRUE)

# sorted data
sort(telephone_data$Bills)
```
If we look at the data, we see several participants that have a first month bill of €0. This does not seem plausible, as the company probably would not give out subscriptions for free. Therefore, we trim the data of zero-values.

```{r}
# row_sub = apply(dd, 1, function(row) all(row !=0 ))
corrected_telephone_data = telephone_data[apply(telephone_data, 1, function(row) all(row !=0 )), ]

par(mfrow=c(2,2))
hist(telephone_data$Bills, xlab = "First month bills (€)", main = "data before correction")
hist(corrected_telephone_data, xlab = "First month bills (€)", main = "data after correction")
boxplot(telephone_data$Bills, ylab = "€")
boxplot(corrected_telephone_data, ylab = "€")
```
We now see that the first bin in the histogram contains a lower frequency due to the removal of zero-values. Based on the boxplots, we see that most participants have relatively low phone bills. Therefore, we might advice the marketing manager to focus on budget subscriptions. 


b) By using a bootstrap test with the test statistic T = median(X1,...,X200), test whether the data telephone.txt stems from the exponential distribution Exp(λ) with some λ from [0.01, 0.1].
```{r}
test_statistic = median(corrected_telephone_data)
lamb = 1/test_statistic

hist(corrected_telephone_data, prob=T, ylim = c(0,0.05))
x = seq(0, length(corrected_telephone_data),length=1000)
f = lamb*exp(-lamb*x )
lines(x,f,col="blue",lwd=2)
```
```{r}
t = median(corrected_telephone_data)
B=1000
tstar=numeric(B)
n=length(corrected_telephone_data)
for (i in 1:B){
  xstar=rexp(n, lamb)
  tstar[i]=median(xstar)}
  hist(tstar,prob=T)
  
p_left=sum(tstar<t)/B
p_right=sum(tstar>t)/B
p=2*min(pl,pr)
pl;pr;p
```

The p-value is .002, therefore, h0 is rejected. This means that the distribution of the data does not come from the exponential distribution with lambda = 1/median(data). 


c) Construct a 95% bootstrap confidence interval for the population median of the sample.

```{r}
B=5000
Tstar=numeric(B)
T1 = median(corrected_telephone_data)

for(i in 1:B) {
  Xstar=sample(corrected_telephone_data,replace=TRUE)
  Tstar[i]=median(Xstar) }

Tstar05=quantile(Tstar,0.05)
Tstar95=quantile(Tstar,0.95)
sum(Tstar<Tstar05)
c(2*T1-Tstar95,2*T1-Tstar05)
```
The 95% bootstrap confidence interval for the population median of the telephone data is [20.7, 36.5] around its median T1=28.9.


d) Assuming X1 , . . . Xn ∼ Exp(λ) and using the central limit theorem for the sample mean, estimate λ and construct again a 95% confidence interval for the population median. Comment on your findings.

We can translate the median of the samples to lambda since
$median[X] = \frac{ln(2)}{\lambda}$
which gives us:
$\lambda = \frac{ln(2)}{median[X]}$

If we plot the medians of the data, we get:
```{r}
hist(Tstar, prob=T)
```

We then check whether the lambdas are normally distributed:
```{r}
shapiro.test(Tstar)

mean_lamb = log(2)/mean(Tstar)
mean_lamb

```



e) Using an appropriate test, test the null hypothesis that the median bill is bigger or equal to 40 euro against the alternative that the median bill is smaller than 40 euro. Next, design and perform a test
to check whether the fraction of the bills less than 10 euro is less than 25%.

```{r}

```



