---
title: "assignment3_ramon"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Exercise 3. Military coups in Africa

To study the influence of different political and geographical variables on the number of military coups, these data are collected for several Sub Saharan African countries in the file africa.txt. The meaning of the different variables:

* miltcoup — number of successful military coups from independence to 1989;
* oligarchy — number years country ruled by military oligarchy from independence to 1989;
* pollib — political liberalization (0 = no civil rights for political expression, 1 = limited civil rights for expression but right to form political parties, 2 = full civil rights);
* parties — number of legal political parties in 1993;
* pctvote — percent voting in last election;
* popn — population in millions in 1989;
* size — area in 1000 square km;
* numelec — total number of legislative and presidential elections;
* numregim — number of regime types.


```{r message=FALSE, warning=FALSE, include=FALSE}
# load data
afr= read.table("africa.txt", header = TRUE)
afr$pollib = as.factor(afr$pollib)
```

a) Perform Poisson regression on the full data set africa, taking miltcoup as response variable. Comment on your findings.
```{r}
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote+popn+size+numelec+numregim, family = poisson, data = afr))
```
We find that oligarchy, pollib and parties contribute the most. In the step-down method below we also end up with a glm based on those three explanatory variables. From the summary above we can see that as oligarchy increases the miltcoups also increase, that when pollib has a higher number the miltcoups decreases, and that when the parties increase the miltcoups increase.

b) Use the step down approach (using output of the function summary) to reduce the number of explanatory variables. Compare the resulting model with your findings in a).

For this step down approach we use the common significance level of 0.05.

Numelec has the highest non-significant p-value, thus we remove that explanatory variable.
```{r}
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote+popn+size+numregim, family = poisson, data = afr))[[12]]
```

numregim has the highest non-significant p-value, thus we remove that explanatory variable.
```{r}
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote+popn+size, family = poisson, data = afr))[[12]]
```

size has the highest non-significant p-value, thus we remove that explanatory variable.
```{r}
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote+popn, family = poisson, data = afr))[[12]]
```

popn has the highest non-significant p-value, thus we remove that explanatory variable.
```{r}
summary(glm(miltcoup~oligarchy+pollib+parties+pctvote, family = poisson, data = afr))[[12]]
```

pctvote has the highest non-significant p-value, thus we remove that explanatory variable.
```{r}
afrglm = glm(miltcoup~oligarchy+pollib+parties, family = poisson, data = afr)
summary(afrglm)
```
With the current model all explanatory variables are significant, no more explanatory variable removal is needed.


The resulting model from the step-down method is as follows:
```
miltcoup = exp(0.207981 + 0.091466*oligarchy -0.495414*pollib1 - 1.112086*pollib2 + 0.022358*parties)
```


```{r}
# plot(predict(afrglm), residuals(afrglm, type="deviance"))
plot(afrglm, 1, ann=FALSE)
title(ylab="Deviance residuals", xlab="Predicted values")
```
The deviance residual's seem to be somewhat explained by the predicted values, therefore the assumptions for this poisson GLM are not met.

Slide 16 of lecture 10 states the following: "Diagnostics for GLM’s is not as straightforward as for linear models, and will not be treated in this course". Because this Poisson regression uses the GLM, no more diagnostics are performed.


c) Predict the number of coups for a hypothetical country for all the three levels of political liberalization and the averages (over all the counties in the data) of all the other (numerical) characteristics. Comment on your findings.

```{r}
# predict(afrglm, data.frame(pollib="0", oligarchy=mean(afr$oligarchy), parties=mean(afr$parties)))[[1]]
# predict(afrglm, data.frame(pollib="1", oligarchy=mean(afr$oligarchy), parties=mean(afr$parties)))[[1]]
# predict(afrglm, data.frame(pollib="2", oligarchy=mean(afr$oligarchy), parties=mean(afr$parties)))[[1]]
```

```{r}
print("pollib 0:")
print(exp(0.207981 + 0.091466*mean(afr$oligarchy) -0.495414*0 - 1.112086*0 + 0.022358*mean(afr$parties)))
print("pollib 1:")
print(exp(0.207981 + 0.091466*mean(afr$oligarchy) -0.495414*1 - 1.112086*0 + 0.022358*mean(afr$parties)))
print("pollib 2:")
print(exp(0.207981 + 0.091466*mean(afr$oligarchy) -0.495414*0 - 1.112086*1 + 0.022358*mean(afr$parties)))
```
Because it predicts a count, we round to the nearest integer. With the given parameters, pollib 0 predicts 2 miltcoups, pollib 1 also predicts 2 miltcoups and pollib 2 predicts 1 miltcoup.